\textbf{Multi-pitch estimation based on mathematical models}\\A pitch estimation framework and harmonic overtone sieve algorithm

### 2021-04-02 paper.tex
stukkie verder

When limiting ourselves to transcribing a guitar, we can use physical limitations of the instrument to our advantage~\cite{physical}. However, this forces the guitarist to use the guitar in a specific way. For instance, these methods fail when the strings are tuned to different notes. Also, this inherently limits the proposed system to guitar transcription, where Fourier based methods are normally instrument invariant.

@@

@@

\subsection{Note sets}  \label{sub:impnoteset}
Using all the found peaks, we can determine which peaks are overtones and which are fundamentals. First, we determine which notes the frequencies represent and calculate the error on these notes. The error is calculated as the number of cents from the closest harmonic note. The errors are different for every overtone (the series of errors is constant for every harmonic overtone series), as overtones are dissonant compared to equal temperament tuning. See table~\ref{tab:overseries} for an example.
\begin{table}[H]\vspace{-3mm}
    \centering
    \begin{tabular}{rrcrr}
        $n$ & $f_{\text{overtone}}$ & closest note & $f_\text{note}$ & error \\
        \hline
        0 & 261.626  & $C_4$    & 261.626  &  0 \\
        1 & 523.251  & $C_5$    & 523.251  &  0 \\
        2 & 784.877  & $G_5$    & 783.991  &  1.955 \\
        3 & 1046.502 & $C_6$    & 1046.502 &  0 \\
        4 & 1308.128 & $E_6$    & 1318.510 &  -13.686 \\
        5 & 1569.753 & $G_6$    & 1567.982 &  1.955 \\
        6 & 1831.379 & $A^\#_6$ & 1864.655 &  -31.174 \\
    \end{tabular}
    \caption{Example of overtone series and the errors compared to the closest note}
    \label{tab:overseries}
\end{table}\vspace{-3mm}
If only harmonic notes were played on a guitar, we could filter all notes with an error higher than an arbitrary threshold. However, as most guitars are not perfectly intonated and strings detune by pressing them down on a fret, this information has to be used with care. Correctly using this information is out of the scope of this paper.

@@

\section{Experiments}  \label{sec:exp}
We will perform a few experiments to test our system. The used audio equipment has a large effect on the results and in turn effects what parameters work best for the system. During development, an electric guitar with active humbucker pickups was used. To convert the analog signal to a sampled digital signal, a Behringer UMC404HD was used. This audio interface has a dynamic range of 100 dB. Like most audio interfaces, this is much less than the dynamic range of a sample (144 dB for 24 bit integer). We opted not to amplify the signal in software, as it will distort it. Also, if amplification is not done carefully, the signal will peak, which means that peaks are cut off.

Often used datasets for evaluating pitch estimation focus on different goals than this paper. For instance, they focus on transcribing multiple instruments at once or having instrument invariance. They also do not focus on real-time transcription, which makes the problem much harder. However, in order to present some comparative measure, we will use the Mirex dataset for "Multiple Fundamental Frequency Estimation \& Tracking"~\cite{data}.

Experiments show basis is there, but results not consistent enough...

@@

\section{Future work}  \label{sec:future}
How to make results more consistent (onset detection, transient filtering). How to make the project more real-time (threading rolling transforms). Also deliberately detuning strings for better results.

@@

>>>>

When limiting ourselves to transcribing a guitar, we can use physical limitations of the instrument to our advantage~\cite{physical}. However, this forces the guitarist to use the guitar in a specific way. For instance, these methods fail when the strings are tuned to different notes.

@@

As stated before, our goal is to perform the transcription in real-time. We choose a constraint of 6 ms. Most musicians work with a driver latency of 9 to 11 ms. On a good computer, this latency can be reduced to 3-5 ms. We will use these 6 ms of latency difference to perform our calculations in. Note that when playing more notes per second, latency becomes more of a problem. On slower pieces, we can get away with much larger latencies. In order to test this for yourself, we developed a tool which plays back the incoming audio signal with a user specifiable delay. This tool can be found under the "latency" branch in our GitHub.

@@

\subsection{Note sets}  \label{sub:impnoteset}
Using all the found peaks, we can determine which peaks are overtones and which are fundamentals. First, we determine which notes the frequencies represent and calculate the error on these notes. The error is calculated as the number of cents from the closest harmonic note. The errors are different for every overtone (the series of errors is constant for every harmonic overtone series), as overtones are dissonant compared to equal temperament tuning. See Table~\ref{tab:overseries} for an example.
\begin{table}[H]\vspace{-3mm}
    \centering
    \begin{tabular}{rrcrr}
        $n$ & $f_{\text{overtone}}$ & closest note & $f_\text{note}$ & error \\
        \hline
        0 & 261.626  & $C_4$    & 261.626  &  0 \\
        1 & 523.251  & $C_5$    & 523.251  &  0 \\
        2 & 784.877  & $G_5$    & 783.991  &  1.955 \\
        3 & 1046.502 & $C_6$    & 1046.502 &  0 \\
        4 & 1308.128 & $E_6$    & 1318.510 &  -13.686 \\
        5 & 1569.753 & $G_6$    & 1567.982 &  1.955 \\
        6 & 1831.379 & $A^\#_6$ & 1864.655 &  -31.174 \\
    \end{tabular}
    \caption{Example of overtone series and the errors compared to the closest note}
    \label{tab:overseries}
\end{table}\vspace{-3mm}
If only harmonic notes were played on a guitar, we could filter all notes with an error higher than an arbitrary threshold. However, as most guitars are not perfectly intonated and strings detune by pressing them down on a fret, this information has to be used with care. Correctly using this information is out of the scope of this paper.

@@

\subsection{UI}
TODO: User interface and user interaction description.

\subsection{Latency}
The latency of our system can be divided in two different latencies, the time to fill a frame with samples and frame processing time. Because our algorithm works on a "per frame" basis instead of a "per sample" basis, we have to wait until the audio driver has fetched enough samples for us to fill a frame. The time it takes to start analysing a frame after receiving the first sample of this frame can be calculated by dividing the number of samples per frame by the sample rate. This is equal to the length of a frame. However,

We use a frame size of 16384 samples and a sample rate of 192000 samples per second, which equates to a frame length of 85.3 ms. This is well over our real-time constraint. Unfortunately, we cannot change this value too much. The sample rate is already at the maximum of most audio interfaces. We also cannot lower the number of samples per frame while using the Fourier transform, as it would further reduce the frequency resolution. However, as will be explained in Section~\ref{sub:futlaten}, this might not be a problem, as we can greatly reduce this latency by employing multi-threading.

We can control the second kind of latency by optimizing our algorithms. In Section~\ref{sub:algspeed}, we will measure the processing time of the different algorithms.

\section{Experiments}  \label{sec:exp}
We will perform a few experiments to test our system. The used audio equipment has a large effect on the results and in turn effects what parameters work best for the system. During development, an electric guitar with active humbucker pickups was used. To convert the analog signal to a sampled digital signal, a Behringer UMC404HD was used. This audio interface has a dynamic range of 100 dB. Like most audio interfaces, this is much less than the dynamic range of a sample (144 dB for 24 bit integer). We opted not to amplify the signal in software, as it will distort it. Also, if amplification is not done carefully, the signal will peak, which means that peaks are cut off. The tests were performed on an AMD 3700X CPU.

Often used datasets for evaluating pitch estimation focus on different goals than this paper. For instance, they focus on transcribing multiple instruments at once or having instrument invariance. They also do not focus on real-time transcription, which makes the problem much harder. However, in order to present some comparative measure, we will use the Mirex dataset for "Multiple Fundamental Frequency Estimation \& Tracking"~\cite{data}. TODO: This dataset consists of actual songs.

We also constructed our own dataset to evaluate the performance. This dataset will be a more technical dataset containing scales and intervals played on a guitar. Scales allow us to measure the monophonic performance. As scales are easy to play, multiple recordings should have the same results. This allows us to isolate and measure other variables, such as the effect of tempo or playing the scale higher/lower. Recordings of different intervals allow us to measure the polyphonic performance. By measuring all two note intervals, we can identify any problems with specific intervals. As the distance in frequency in an interval increases when the same interval is played at higher notes, we record the intervals at multiple root notes. We will also measure the performance on the four basic triads: the major triad, minor triad, augmented triad and diminished triad.

\subsection{Latency}  \label{sub:algspeed}
The processing time of our system is very important, as it will directly impact the latency.

@@

\section{Future work}  \label{sec:future}
How to make results more consistent (onset detection, transient filtering). How to make the project more real-time (threading rolling transforms). Also deliberately detuning strings for better results. Amplification->more overtones detected but also more noise.

@@

\subsection{Multi-threading}  \label{sub:futlaten}
asdf
